A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-10 paper/paper.tex
Add more to the intro

The goal of this paper is to give a tutorial on proof by logical relation for
dependent types in a simple and digestable format. % The language specification will
% be given in Section~\ref{sec:spec}.
The system we present is small but not minimal. We want to cover
features that would require more interesting treatments under
dependent types without introducing complexities that would have
existed in its simply typed variant. We choose boolean
types over natural numbers as our base type for simplicity, but include an
intensional identity type in our type system as the interpretation
of identity types requires extra side conditions that are not present
in other types. Unlike \citet{nbeincoq,decagda,martin-lof-a-la-coq}
but similar to \citet{anand2014towards}, we include an infinite
hierarchy of universes since we want to not only have at least one
predicative universe level to support type-level computations, but
also avoid the unnecessary code duplication pointed out by
\citet{nbeincoq}.
We choose consistency as the property of interest since it is the one
of the easiest metatheoretic property that requires proof by logical
relation.

@@

In Section~\ref{sec:logrelmech}, we discuss the details about our
mechanization, including our use of existing libraries such as
Autosubst for handling bindings and CoqHammer for general-purpose
automation. In Section~\ref{sec:discuss}, we show that extensions such
as $\eta$ rules for functions and normalization for open terms are
more or less orthogonal to the fact that the language is dependently
typed. In our mechanized proof of said extensions, the overall
structure of the logical relation is preserved, and the additional
proof obligations are mostly related to properties about the untyped
lambda calculus, all of which can be derived through syntactic
means.
In other words, once we have established the base technique for
proof by logical relation for dependent types, we can factor out the
complexities of an extension the are not specifically related to dependent
types; such an extension, if desired, may be studied in the context of
a simply typed language and later ported into a dependently typed setting.

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. To help
the reader understand the binding structure of the terms, we annotate
the terms with their DeBruijn depth. For example, ....
 We use
the unscoped DeBruijn representation to leave less ambiguity for the
formulation of the system and simplify the reasoning about binders.
Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
syntax, including renaming, substitution, and operations over
the typing context or substitution.
As a
dependent type theory, terms and types are collapsed into the same
syntactic category. Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable. $[[Set i]]$ represents the universe
type where $[[i]]$ ranges over the set of natural numbers.
 Finally,
\lang also includes an intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.

@@

Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$.
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests that it is an equivalence relation, and in fact, we
will use coherence as the equational theory for our type conversion
rule. Now, we prove that coherence is indeed an equivalence relation.

>>>>

The goal of this paper is to give a tutorial on proof by logical relation for
dependent types in a simple and digestable format. % The language specification will
% be given in Section~\ref{sec:spec}.
The system we present is small but not minimal. We want to cover
features that are only available or require special treatments under
dependent types%  without introducing complexities that would have
% existed in its simply typed variant
. We choose boolean
types over natural numbers as our base type for simplicity, but include an
intensional identity type in our type system as the interpretation
of identity types requires an extra side condition that is needed for
indexed family of types. Unlike \citet{nbeincoq,decagda,martin-lof-a-la-coq}
but similar to \citet{anand2014towards}, we include an infinite
hierarchy of universes since we want to not only have at least one
predicative universe level to support type-level computations, but
also avoid the unnecessary code duplication pointed out by
\citet{nbeincoq}.

@@

In Section~\ref{sec:logrelmech}, we discuss the details about our
mechanization, including our use of existing libraries such as
Autosubst for handling bindings and CoqHammer for general-purpose
automation. In Section~\ref{sec:discuss}, we extend our logical
relation to prove the stronger property that every well-typed term
has a normal form. Most interestingly, we show that extensions such
as $\eta$ rules for functions and normalization for open terms are
mostly orthogonal to the fact that the language is dependently
typed. In our mechanized proof, the overall
structure of the logical relation remains unchanged, and the additional
proof obligations are mostly related to properties about the untyped
lambda calculus, all of which can be derived through syntactic
means.
In other words, once we have established the base technique for
proof by logical relation for dependent types, we can factor out the
complexities of an extension the are not specifically related to dependent
types; such an extension, if desired, may be studied in the context of
a simply typed language and later ported into a dependently typed setting.

@@

The syntax of \lang can be found in Figure~\ref{fig:syntax}. We will use
the unscoped de Bruijn representation for the rest of the paper,
adopting notations from \citet{autosubst2}, summarized in Figure~\ref{fig:auxdef}.
Without providing
the full definition of the renaming and substitution functions, it is
impossible to tell the binding structure from the syntax
alone. Therefore, we annotate the syntax in Figure~\ref{fig:syntax}
with the de Bruijn depth of each term, though we note that the
syntax we work with is unscoped and the choice does matter when
we extend our logical relation to open terms in Section~\ref{sec:discuss}.

% Figure~\ref{fig:auxdef} shows the auxiliary definitions over the term
% syntax, including renaming, substitution, and operations over
% the typing context or substitution.
As a
dependent type theory, terms and types are collapsed into the same
syntactic category. Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable. $[[Set i]]$ represents the universe
type where $[[i]]$ ranges over the set of natural numbers.
 Finally,
\lang also includes an intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.

@@

Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$.
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests that it is an equivalence relation, and in fact, we
will use coherence as the equational theory for our type conversion
rule.

We sketch out some key properties about coherence without giving their
proofs. Our technique for establishing those results is based on
\citet{takahashi-parallel-reduction} and a modern exposition can be
found in \citet{plfa22.08}.

Now, we prove that coherence is indeed an equivalence relation.
